---
title: "Chapter 6"
subtitle: "Maps"
author: "Aditya Dahiya"
date: 2023-11-18
format: 
  html: 
    code-fold: true
    code-copy: hover
    code-link: true
execute: 
  echo: true
  warning: false
  error: false
  cache: true
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/ggplot2book3e/Chapter5.html"
  description: "Solutions Manual (and Beyond) for ggplot2: Elegant Graphics for Data Analysis (3e)"
  twitter: true
  facebook: true
  linkedin: true
  email: true
  mastodon: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

::: callout-note
This Chapter does not have Exercises. The code and examples below show map plotting in `R` with `tidyverse` and `sf` with examples from India.
:::

```{r}
#| label: setup

library(tidyverse)     # everything and ggplot2
library(sf)            # sf and shape files
library(ggthemes)      # for theme_map
library(ggspatial)     # map annotations

# Reading in the shape file of the map of India
india_map <- read_sf(here::here("data", 
                                "india_map", 
                                "India_State_Boundary.shp")) |> 
  # tidy names of variables
  rename(state = State_Name) |> 
  # Renaming to official names. Official names taken from Government:
  # https://knowindia.india.gov.in/states-uts/
  mutate(state = case_when(
    state == "Andaman & Nicobar" ~ "Andaman and Nicobar Islands",
    state == "Daman and Diu and Dadra and Nagar Haveli" ~ "Dadra and Nagar Haveli and Daman & Diu",
    state == "Jammu and Kashmir" ~ "Jammu & Kashmir",
    state == "Telengana" ~ "Telangana",
    .default = state
  ))

# Add names of Union Territories
union_territories <- c(
  "Andaman and Nicobar Islands",
  "Chandigarh",
  "Dadra and Nagar Haveli and Daman & Diu",
  "Delhi",
  "Jammu & Kashmir",
  "Ladakh",
  "Lakshadweep",
  "Puducherry"
)

# Getting in a dataframe with map of State of Haryana in India
haryana_map <- read_sf(here::here("data",
                                  "haryana_map",
                                  "HARYANA_DISTRICT_BDY.shp")) |> 
  janitor::clean_names() |> 
  mutate(
    district = str_replace_all(district,
                               pattern = ">",
                               replacement = "A"),
    state = str_replace_all(state,
                            pattern = ">",
                            replacement = "A"),
    district = case_when(
      district == "FAR|DABAD" ~ "FARIDABAD",
      district == "J|ND" ~ "JIND",
      district == "PAN|PAT" ~ "PANIPAT",
      district == "SON|PAT" ~ "SONIPAT",
      .default = district
    ),
    district = snakecase::to_title_case(district)
      )

```

## **6.2 Simple features maps**

First, we download a shape file of India's map (from [here](https://github.com/AnujTiwari/India-State-and-Country-Shapefile-Updated-Jan-2020)) and plot a map of India with latest state and Union Territory boundaries. The code below shows a simple example of ease of plotting with `ggplot()` and `geom_sf()` along with **labeling** [official names](https://knowindia.india.gov.in/states-uts/) of States and Union Territories.

```{r}
#| fig-width: 10
#| fig-asp: 1

# Plotting the actual map
india_map |> 
  
  # A variable to use for different font colour for States vs. Union Territories
  mutate(type = if_else(state %in% union_territories,
                        "Union Territory",
                        "State")) |> 
  # Start Plot
  ggplot(aes(geometry = geometry,
             col = type)) + 
  geom_sf(col = "darkgrey", fill = "white") +
  geom_sf_text(aes(label = state,
                size = type)) +
  
  # Colour and Size Scales
  scale_color_manual(values = c("darkblue", "darkred")) +
  scale_size_discrete(range = c(3.5, 2.5)) +
  
  # Themes
  theme_map() +
  theme(legend.position = "bottom") +
  labs(col = NULL) +
  
  # Adding Scale and North Arrow
  annotation_scale(bar_cols = c("darkgrey", "white"),
                   location = "br") +
  annotation_north_arrow(location = "tr", 
                         which_north = "true") +
  guides(size = "none")
```

## **6.2.1 Layered maps and 6.2.2 Labelled maps**

We can use data from multiple data-frames and add them as layers to a single map using `ggplot2` as shown below: --

```{r}

# Names of Bordering states of Haryana
bordering_states = c("Punjab",
                     "Delhi",
                     "Himachal Pradesh",
                     "Rajasthan",
                     "Uttar Pradesh",
                     "Uttarakhand")

# Start plotting with India Map showing only bordering states
ggplot(data = india_map |> 
         filter(state %in% bordering_states),
       mapping = aes(geometry = geometry,
                     label = state)) +
  
  geom_sf(fill = "white",
          col = "darkgrey",
          alpha = 0.5) +
  geom_sf_text(col = "darkgrey") +
  
  # Map of Haryana with Districts
  geom_sf(data = haryana_map,
          fill = "lightgrey",
          col = "black") +
  
  # Labels and theme
  labs(x = NULL, y = NULL,
       title = "Combining two sf data-frames into a single map") +
  theme_minimal()
```

## **6.2.3 Adding other geoms**

Plotting a specific state from i..e, Haryana, and its districts along with their area (in thousand sq. km.) using official data from [Survey of India](https://onlinemaps.surveyofindia.gov.in/), and adding geoms from other metadata - area and length of district borders: --

```{r}
#| fig-width: 10
#| fig-asp: 1

haryana_map |>
  select(-c(remarks, state_lgd, district_l)) |> 
  arrange(desc(shape_area)) |> 
  mutate(
    rank = row_number(),
    area = round(shape_area/1e9, 2),
    font_col = district %in% c("Hisar", "Sirsa")
  ) |> 
  
  ggplot(aes(geometry = geometry,
             label = paste0(district),
             fill = shape_area/1e6)) +
  geom_sf(col = "white") +
  geom_sf_text(aes(col = font_col),
               size = 4) +
  theme_map() +
  annotation_scale(location = "br") +
  annotation_north_arrow(location = "tr") +
  labs(fill = "District Area\n(in sq. km.)",
       title = "Haryana: Districts (with area)") +
  scale_fill_viridis_c(labels = scales::label_comma(),
                       option = "G") +
  scale_color_manual(values = c("white", "black")) +
  guides(col = "none")
```

### Chloropleth with `tmap`

A package to create chloropleths very easily is the `tmap` package [@tmap] of `R` . However, is uses a very different syntax than `ggplot2` . An interesting feature is the ability to plot interactive maps. However, since it is outside `ggplot2` grammar and syntax, I do not use it here.

```{r}
#| eval: false

library(tmap)
tm_hy <- haryana_map |> 
  mutate(area = shape_area/1e6)

t_hy <- tm_shape(tm_hy) +
  tm_polygons(col = "area",
              title = "District Area\n(in sq. km.)",
              border.col = "black",
              interactive = TRUE,
              style = "pretty")
t_hy
tmap_mode("view")
tmap_last()

tmap_mode("plot")
```

### Adding `leaflet` maps to Chloropleths

```{r}
library(leaflet)

# Create a quantile of colours' palette to be used
pal_hy <- colorQuantile("Blues", 
                        domain = NULL,
                        n = 5)

# Create vector of text to display on pop-ups in leaflet map
p_popup <- paste0(haryana_map$district, 
                  " District.  Area: ", 
                  round(haryana_map$shape_area/1e6, 0),
                  " sq. km.")

# Create leaflet map
# Data set
haryana_map |> 
  
  # Transform polygons into CRS=4326 since leaflet only understand that
  st_transform(crs = 4326) |> 
  
  # Begin leaflet map
  leaflet() |> 
  
  # Add polygons from the geometry column of the data-set
  addPolygons(
    stroke = FALSE,                       # remove polygon borders
    fillColor = ~ pal_hy(shape_area/1e6), # set fill color with function
    fillOpacity = 0.6,                    # translucent to see background map
    smoothFactor = 0.5,                   # make it nicer
    popup = p_popup,                      # add popup
    group = "District Area"               # a Group label for leaflet options
  ) |> 
  
  # Add base map from leaflet; default is Open Street Maps
  addTiles() |>                           # Adding Base Map
  
  # Adding a legend
  addLegend(
    position = "bottomright",  # location
    pal = pal_hy,              # palette function
    values = ~shape_area/1e6,  # value to be passed to palette function
    title = "District Area (sq. km.)" # legend title
    ) |> 
  
  # Adding an option to view different base maps
  addLayersControl(
    baseGroups = c("OSM", "Carto"),
    overlayGroups = c("District Area")
  )
```

## **6.4 Working with `sf` data**

We can even drawn out different pieces of polygons, i.e. islands, enclaves or exclaves using `sf` data. Here's an example for India's island chains: --

```{r}
#| layout-ncol: 3
#| fig-asp: 3 

# Pull out the geometry of Andaman and Nicobar Islands as "ani"
ani <- india_map |> 
  filter(state == "Andaman and Nicobar Islands") |> 
  pull(geometry)

# Bounding Box of ani
lims <- st_bbox(ani)

ani_islands <- st_cast(ani, "POLYGON")

# Checking order: We see that islands are listed in decreasing order of size
# order(st_area(ani_islands), decreasing = TRUE)

# A common theme to use in all Island Maps
theme_islands <- theme_map() +
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5,
                                   size = 25),
        axis.text.y = element_text(vjust = 0.5,
                                   size = 25),
        plot.title.position = "plot",
        panel.background = element_rect(fill = "lightblue"),
        plot.title = element_text(size = 25))

# Drawing the complete Andaman and Nicobar Islands
india_map |> 
  filter(state == "Andaman and Nicobar Islands") |> 
  ggplot() +
  geom_sf(fill = "white") +
  labs(title = "Andaman and Nicobar Islands") +
  scale_x_continuous(limits = c(lims["xmin"], lims["xmax"])) +
  scale_y_continuous(limits = c(lims["ymin"], lims["ymax"])) +
  theme_islands

# Drawing the largest of the Andaman and Nicobar Islands
ggplot(ani_islands[1]) +
  geom_sf(fill = "white") +
  labs(title = "Largest Island: Middle Andaman") +
  scale_x_continuous(limits = c(lims["xmin"], lims["xmax"])) +
  scale_y_continuous(limits = c(lims["ymin"], lims["ymax"])) +
  theme_islands

# Drawing the second largest of the Andaman and Nicobar Islands
ggplot(ani_islands[2]) +
  geom_sf(fill = "white") +
  labs(title = "2nd Largest: North Andaman") +
  scale_x_continuous(limits = c(lims["xmin"], lims["xmax"])) +
  scale_y_continuous(limits = c(lims["ymin"], lims["ymax"])) +
  theme_islands
```

## 6.6 Data Sources

The official source for maps of India, shape files is [here](https://onlinemaps.surveyofindia.gov.in/Digital_Product_Show.aspx)
