---
title: "Chapter 18"
subtitle: "Programming with ggplot2"
author: "Aditya Dahiya"
date: "2024-03-31"
format: 
  html: 
    code-fold: true
    code-copy: hover
    code-link: true
execute: 
  echo: true
  warning: false
  error: false
  cache: true
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/ggplot2book3e/Chapter18.html"
  description: "Solutions Manual (and Beyond) for ggplot2: Elegant Graphics for Data Analysis (3e)"
  twitter: true
  facebook: true
  linkedin: true
  email: true
  mastodon: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r}
#| label: setup

library(tidyverse)
library(ggthemes)
library(scales)
library(cowplot)
library(colorspace)
```

## 18.2.1 Exercises

### Question 1

**Create an object that represents a pink histogram with 100 bins.**

In the following code, `pink_hist` is an object that represents a pink histogram with 100 bins. This object can be added to any `ggplot()` function call.

```{r}
#| code-fold: false
#| echo: true

pink_hist <- geom_histogram(
    bins = 100,
    colour = "#5A0020",
    fill = "pink",
    linewidth = 0.5
  )
```

We can use this object `pink_hist` in the following way to produce @fig-ex2-q1

```{r}
#| label: fig-ex2-q1
#| fig-cap: "An object that represents a pink histogram with 100 bins"
#| code-fold: true

df <- tibble(
  x = rnorm(n = 1000, mean = 50, sd = 10),
  y = x %/% 10
  )

df |> 
  ggplot(aes(x)) +
  theme_cowplot() +
  pink_hist
```

### Question 2

**Create an object that represents a fill scale with the Blues ColorBrewer palette.**

The object `blues_fill` represents a scale using `ColorBrewer` palette "Blues", which is a discrete palette of 9 different colours.

```{r}
#| code-fold: false
#| eval: true
#| echo: true

blues_fill <- scale_fill_brewer(palette = "Blues")
```

The @fig-ex2-q2 uses the `blues_fill` to generate a colored histogram.

```{r}
#| label: fig-ex2-q2

ggplot(df, aes(x, fill = as_factor(y), group = y)) +
  geom_histogram(bins = 100) +
  theme_clean() +
  theme(legend.position = "none") +
  blues_fill
```

### Question 3

**Read the source code for `theme_grey()`. What are its arguments? How does it work?**

The `theme_grey()` is a function that sets the visual theme for a ggplot object to a grey background with white gridlines. Here's a summary of its arguments and how it works. **Internally, its code creates a theme object `t` , which it then adds to a ggplot2 plot to generate the desired default theme plot.** The folded code shows the actual source code and the table below shows the purpose of each argument: —

```{r}
#| code-fold: true
#| eval: false


theme(
  line = element_line(
    colour = "black", 
    linewidth = base_line_size, 
    linetype = 1, 
    lineend = "butt"
    ), 
  rect = element_rect(
    fill = "white", 
    colour = "black", 
    linewidth = base_rect_size, 
    linetype = 1
    ), 
  text = element_text(
    family = base_family, 
    face = "plain", 
    colour = "black", 
    size = base_size, 
    lineheight = 0.9, 
    hjust = 0.5, 
    vjust = 0.5, 
    angle = 0, 
    margin = margin(), 
    debug = FALSE), 
  axis.line = element_blank(), 
  axis.line.x = NULL, 
  axis.line.y = NULL, 
  axis.text = element_text(
    size = rel(0.8),
    colour = "grey30"), 
  axis.text.x = element_text(
    margin = margin(t = 0.8 * half_line/2), 
    vjust = 1
    ), 
  axis.text.x.top = element_text(
    margin = margin(b = 0.8 * half_line/2), 
    vjust = 0
    ), 
  axis.text.y = element_text(
    margin = margin(r = 0.8 * half_line/2), 
    hjust = 1
    ), 
  axis.text.y.right = element_text(
    margin = margin(l = 0.8 * half_line/2), 
    hjust = 0
    ), 
  axis.text.r = element_text(
    margin = margin(l = 0.8 * half_line/2, r = 0.8 * half_line/2),
    hjust = 0.5),
  axis.ticks = element_line(colour = "grey20"), 
  axis.ticks.length = unit(half_line/2, "pt"), 
  axis.ticks.length.x = NULL, 
  axis.ticks.length.x.top = NULL, 
  axis.ticks.length.x.bottom = NULL, 
  axis.ticks.length.y = NULL, 
  axis.ticks.length.y.left = NULL, 
  axis.ticks.length.y.right = NULL, 
  axis.minor.ticks.length = rel(0.75), 
  axis.title.x = element_text(
    margin = margin(t = half_line/2), 
    vjust = 1), 
  axis.title.x.top = element_text(
    margin = margin(b = half_line/2), 
    vjust = 0
    ), 
  axis.title.y = element_text(
    angle = 90,
    margin = margin(r = half_line/2), 
    vjust = 1
    ), 
  axis.title.y.right = element_text(
    angle = -90,
    margin = margin(l = half_line/2), 
    vjust = 1
    ), 
  legend.background = element_rect(colour = NA), 
  legend.spacing = unit(2 * half_line, "pt"), 
  legend.spacing.x = NULL, 
  legend.spacing.y = NULL, 
  legend.margin = margin(
    half_line, 
    half_line, 
    half_line, 
    half_line), 
  legend.key = NULL,
  legend.key.size = unit(1.2, "lines"), 
  legend.key.height = NULL, 
  legend.key.width = NULL, 
  legend.key.spacing = unit(half_line, "pt"), 
  legend.text = element_text(size = rel(0.8)), 
  legend.title = element_text(hjust = 0), 
  legend.ticks.length = rel(0.2), 
  legend.position = "right", 
  legend.direction = NULL, 
  legend.justification = "center", 
  legend.box = NULL, 
  legend.box.margin = margin(0, 0, 0, 0, "cm"), 
  legend.box.background = element_blank(), 
  legend.box.spacing = unit(2 * half_line, "pt"), 
  panel.background = element_rect(fill = "grey92", colour = NA), 
  panel.border = element_blank(), 
  panel.grid = element_line(colour = "white"), 
  panel.grid.minor = element_line(linewidth = rel(0.5)), 
  panel.spacing = unit(half_line, "pt"), panel.spacing.x = NULL, 
  panel.spacing.y = NULL, 
  panel.ontop = FALSE, 
  strip.background = element_rect(fill = "grey85", colour = NA), 
  strip.clip = "inherit", 
  strip.text = element_text(
    colour = "grey10", 
    size = rel(0.8), 
    margin = margin(
      0.8 * half_line, 
      0.8 * half_line, 
      0.8 * half_line, 
      0.8 * half_line)
    ), 
  strip.text.x = NULL, 
  strip.text.y = element_text(angle = -90), 
  strip.text.y.left = element_text(angle = 90), 
  strip.placement = "inside",
  strip.placement.x = NULL, 
  strip.placement.y = NULL, 
  strip.switch.pad.grid = unit(half_line/2, "pt"), 
  strip.switch.pad.wrap = unit(half_line/2, "pt"), 
  plot.background = element_rect(fill = "white"), 
  plot.title = element_text(
    size = rel(1.2), 
    hjust = 0, 
    vjust = 1, 
    margin = margin(b = half_line)
    ), 
  plot.title.position = "panel",
  plot.subtitle = element_text(
    hjust = 0, 
    vjust = 1, 
    margin = margin(b = half_line)
    ), 
  plot.caption = element_text(
    size = rel(0.8), 
    hjust = 1,
    vjust = 1, 
    margin = margin(t = half_line)
    ), 
  plot.caption.position = "panel",
  plot.tag = element_text(
    size = rel(1.2),
    hjust = 0.5, 
    vjust = 0.5
    ), 
  plot.tag.position = "topleft", 
  plot.margin = margin(
    half_line, 
    half_line, 
    half_line, 
    half_line
    ), 
  complete = TRUE
  )
```

**Arguments and their purpose:**

| Argument to base `theme()` call                                                                                                                                                          | Meaning and Purpose                                                                                                                                                                         |
|------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| `line = element_line(colour = "black", linewidth = base_line_size, linetype = 1, lineend = "butt")`                                                                                      | Specifies the appearance of lines in the plot, setting the color to black, the line width to the base line size, the line type to solid, and the line end style to "butt".                  |
| `rect = element_rect(fill = "white", colour = "black", linewidth = base_rect_size, linetype = 1)`                                                                                        | Sets the appearance of rectangles in the plot, defining the fill color as white, the border color as black, the border width as the base rectangle size, and the border line type as solid. |
| `text = element_text(family = base_family, face = "plain", colour = "black", size = base_size, lineheight = 0.9, hjust = 0.5, vjust = 0.5, angle = 0, margin = margin(), debug = FALSE)` | Defines the appearance of text in the plot, specifying the font family, face, color, size, line height, horizontal and vertical justification, angle, margin, and debug options.            |
| `axis.line = element_blank()`                                                                                                                                                            | Removes the axis lines from the plot.                                                                                                                                                       |
| `axis.line.x = NULL`                                                                                                                                                                     | No horizontal axis line is specified.                                                                                                                                                       |
| `axis.line.y = NULL`                                                                                                                                                                     | No vertical axis line is specified.                                                                                                                                                         |
| `axis.text = element_text(size = rel(0.8), colour = "grey30")`                                                                                                                           | Sets the appearance of axis text, defining the size relative to the base size and the color as grey30.                                                                                      |
| `axis.text.x = element_text(margin = margin(t = 0.8 * half_line/2), vjust = 1)`                                                                                                          | Specifies the appearance of horizontal axis text, setting the margin and vertical justification.                                                                                            |
| `axis.text.x.top = element_text(margin = margin(b = 0.8 * half_line/2), vjust = 0)`                                                                                                      | Specifies the appearance of top horizontal axis text, setting the margin and vertical justification.                                                                                        |
| `axis.text.y = element_text(margin = margin(r = 0.8 * half_line/2), hjust = 1)`                                                                                                          | Defines the appearance of vertical axis text, setting the margin and horizontal justification.                                                                                              |
| `axis.text.y.right = element_text(margin = margin(l = 0.8 * half_line/2), hjust = 0)`                                                                                                    | Specifies the appearance of right vertical axis text, setting the margin and horizontal justification.                                                                                      |
| `axis.text.r = element_text(margin = margin(l = 0.8 * half_line/2, r = 0.8 * half_line/2), hjust = 0.5)`                                                                                 | Defines the appearance of reverse vertical axis text, setting the margin and horizontal justification.                                                                                      |
| `axis.ticks = element_line(colour = "grey20")`                                                                                                                                           | Sets the appearance of axis ticks, defining the color as grey20.                                                                                                                            |
| `axis.ticks.length = unit(half_line/2, "pt")`                                                                                                                                            | Specifies the length of axis ticks as half the line size.                                                                                                                                   |
| `axis.minor.ticks.length = rel(0.75)`                                                                                                                                                    | Sets the length of minor axis ticks relative to the base size.                                                                                                                              |
| `axis.title.x = element_text(margin = margin(t = half_line/2), vjust = 1)`                                                                                                               | Defines the appearance of horizontal axis title, setting the margin and vertical justification.                                                                                             |
| `axis.title.x.top = element_text(margin = margin(b = half_line/2), vjust = 0)`                                                                                                           | Specifies the appearance of top horizontal axis title, setting the margin and vertical justification.                                                                                       |
| `axis.title.y = element_text(angle = 90, margin = margin(r = half_line/2), vjust = 1)`                                                                                                   | Sets the appearance of vertical axis title, defining the angle, margin, and vertical justification.                                                                                         |
| `axis.title.y.right = element_text(angle = -90, margin = margin(l = half_line/2), vjust = 1)`                                                                                            | Specifies the appearance of right vertical axis title, defining the angle, margin, and vertical justification.                                                                              |
| `legend.background = element_rect(color = NA)`                                                                                                                                           | Sets the legend background appearance with no border color.                                                                                                                                 |
| `legend.spacing = unit(2 * half_line, "pt")`                                                                                                                                             | Specifies the spacing between legend elements as twice the line size.                                                                                                                       |
| `legend.margin = margin(half_line, half_line, half_line, half_line)`                                                                                                                     | Sets the margin around the legend.                                                                                                                                                          |
| `panel.background = element_rect(fill = "grey92", colour = NA)`                                                                                                                          | Defines the appearance of the panel background, setting the fill color as grey92 and removing the border color.                                                                             |
| `panel.grid = element_line(colour = "white")`                                                                                                                                            | Specifies the appearance of panel grid lines, setting the color to white.                                                                                                                   |
| `panel.grid.minor = element_line(linewidth = rel(0.5))`                                                                                                                                  | Sets the appearance of minor panel grid lines, defining the line width relative to the base size.                                                                                           |
| `panel.spacing = unit(half_line, "pt")`                                                                                                                                                  | Specifies the spacing between panels as the half line size.                                                                                                                                 |
| `strip.background = element_rect(fill = "grey85", colour = NA)`                                                                                                                          | Sets the appearance of strip background, defining the fill color as grey85 and removing the border color.                                                                                   |
| `strip.text = element_text(colour = "grey10", size = rel(0.8), margin = margin(0.8 * half_line, 0.8 * half_line, 0.8 * half_line, 0.8 * half_line))`                                     | Defines the appearance of strip text, setting the color, size, and margin.                                                                                                                  |
| `strip.text.y = element_text(angle = -90)`                                                                                                                                               | Specifies the appearance of vertical strip text, defining the angle.                                                                                                                        |
| `strip.text.y.left = element_text(angle = 90)`                                                                                                                                           | Sets the appearance of left vertical strip text, defining the angle.                                                                                                                        |
| `plot.background = element_rect(colour = "white")`                                                                                                                                       | Defines the appearance of plot background, setting the border color as white.                                                                                                               |
| `plot.title = element_text(size = rel(1.2), hjust = 0, vjust = 1, margin = margin(b = half_line))`                                                                                       | Specifies the appearance of plot title, setting the size, horizontal and vertical justification, and margin.                                                                                |
| `plot.subtitle = element_text(hjust = 0, vjust = 1, margin = margin(b = half_line))`                                                                                                     | Sets the appearance of plot subtitle, defining the horizontal and vertical justification, and margin.                                                                                       |
| `plot.caption = element_text(size = rel(0.8), hjust = 1, vjust = 1, margin = margin(t = half_line))`                                                                                     | Defines the appearance of plot caption, setting the size, horizontal and vertical justification, and margin.                                                                                |
| `plot.tag = element_text(size = rel(1.2), hjust = 0.5, vjust = 0.5)`                                                                                                                     | Specifies the appearance of plot tag, setting the size, horizontal and vertical justification.                                                                                              |
| `plot.margin = margin(half_line, half_line, half_line, half_line)`                                                                                                                       | Sets the margin around the plot.                                                                                                                                                            |

**How it works:**

`theme_grey()` creates a set of **nice, elegant and default theme settings** for ggplot objects that use a grey background with white gridlines. It sets the base sizes, colors, and shapes for various plot elements such as text, lines, rectangles, and points. It also allows customization of line end styles, line join styles, and line mitre limits.

The function returns a `theme` object that can be applied to ggplot objects using the `theme()` function. When applied, this theme modifies the visual appearance of the plot elements according to the settings specified in `theme_grey()`.

### Question 4

**Create `scale_colour_wesanderson()`. It should have a parameter to pick the palette from the `wesanderson` package, and create either a continuous or discrete scale.**

The following code creates a `scale_colour_wesanderson()` and `scale_fill_wesanderson()` functions for this purpose.

```{r}
#| code-fold: false
#| echo: true
#| eval: true

scale_colour_wesanderson <- function(palette, type = "continuous", ...) {
  if (type == "continuous") {
    return(
      scale_colour_gradientn(
        colours = wesanderson::wes_palette(palette),
        ...
      )
    )
  } else if (type == "discrete") {
    return(
      scale_colour_manual(
        values = wesanderson::wes_palette(palette),
        ...
      )
    )
  } else {
    stop("Unsupported scale type. Use 'continuous' or 'discrete'.")
  }
}


scale_fill_wesanderson <- function(palette, type = "continuous", ...) {
  if (type == "continuous") {
    return(
      scale_fill_gradientn(
        colours = wesanderson::wes_palette(palette),
        ...
      )
    )
  } else if (type == "discrete") {
    return(
      scale_fill_manual(
        values = wesanderson::wes_palette(palette),
        ...
      )
    )
  } else {
    stop("Unsupported scale type. Use 'continuous' or 'discrete'.")
  }
}
```

The @fig-ex2-q4-2 demonstrates the use of this `scale_fill_wesanderson()` function using a dataset `heatmap` from **{wesanderson}** package itself. [@wesanderson].

```{r}
#| label: fig-ex2-q4
#| fig-cap: "Demonstrating the use of our custom created function scale_fill_wesanderson()"
#| fig-subcap: 
#|   - "Base plot"
#|   - "Upon adding scale_fill_wesanderson() to base plot"

q4_1 <- wesanderson::heatmap |> 
  ggplot(aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  theme_void()

q4_1
q4_1 +
  scale_fill_wesanderson(
    palette = "Zissou1",
    type = "continuous"
    )
```
