---
title: "Chapter 21"
subtitle: "A case study"
author: "Aditya Dahiya"
date: "2024-09-29"
format: 
  html: 
    code-fold: true
    code-copy: hover
    code-link: true
    mermaid:
      theme: neutral
execute: 
  echo: true
  warning: false
  error: false
  cache: true
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/ggplot2book3e/Chapter21.html"
  description: "Solutions Manual (and Beyond) for ggplot2: Elegant Graphics for Data Analysis (3e)"
  twitter: true
  facebook: true
  linkedin: true
  email: true
  mastodon: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

::: callout-warning
There are no exercises in this chapter. Instead, I use the code and teachings of this chapter to make a custom `ggplot2` extension.
:::

## Creating a new ggplot2 extension: `geom_wave()`

The provided code defines a ggplot2 extension which allows users to create **wave-like geoms.** These set of functions and geoms to create and customize wave patterns. The main functions are:

-   `geom_wave()`: Creates a wave geom.

-   `stat_wave()`: Computes wave paths for a given dataset.

-   `scale_amplitude_continuous()` and `scale_frequency_continuous()`: Scales for amplitude and frequency of the drawn waves.

The complete code: —

```{r}
#| label: code-for-geom-wave

# Loading libaries
library(tidyverse)
library(scales)
library(grid)
library(gt)

# Defining a wave creating function
create_wave <- function(x,
                        y,
                        xend,
                        yend,
                        amplitude = 1,
                        frequency = 1,
                        n = 5000) {
  # Validate the input arguments
  if (amplitude < 0) {
    rlang::abort("`amplitude` must be non-negative.")
  }
  if (frequency <= 0) {
    rlang::abort("`frequency` must be larger than zero.")
  }
  if (n == 0) {
    rlang::abort("`n` must be greater than zero.")
  }

  # Calculate the direct length of the wave path
  length <- sqrt((x - xend)^2 + (y - yend)^2)

  # Calculate the number of points we need
  n_points <- n

  # Calculate the sequence of x and y offset values
  x_seq <- seq(x, xend, length.out = n_points)
  y_seq <- seq(y, yend, length.out = n_points)

  # Create and return the transformed data frame
  data.frame(
    x = x_seq,
    y = y_seq + sin((x_seq - x) * frequency / (length * pi * 2)) * amplitude
  )
}

# A wave Grob
waveGrob <- function(x0 = unit(0, "npc"),
                     y0 = unit(0, "npc"),
                     x1 = unit(1, "npc"),
                     y1 = unit(1, "npc"),
                     amplitude = unit(0.1, "npc"),
                     frequency = 1,
                     n = 5000,
                     default.units = "npc",
                     name = NULL,
                     gp = gpar(),
                     vp = NULL) {
  
  # Use the default unit if the user does not specify one
  if (!is.unit(x0)) x0 <- unit(x0, default.units)
  if (!is.unit(x1)) x1 <- unit(x1, default.units)
  if (!is.unit(y0)) y0 <- unit(y0, default.units)
  if (!is.unit(y1)) y1 <- unit(y1, default.units)
  if (!is.unit(amplitude)) amplitude <- unit(amplitude, default.units)

  # Return a gTree of class "wave"
  gTree(
    x0 = x0,
    y0 = y0,
    x1 = x1,
    y1 = y1,
    amplitude = amplitude,
    frequency = frequency,
    n = n,
    name = name,
    gp = gp,
    vp = vp,
    cl = "wave"
  )
}

makeContent.wave <- function(x) {
  
  # Convert position and amplitude values to absolute units
  x0 <- convertX(x$x0, "mm", valueOnly = TRUE)
  x1 <- convertX(x$x1, "mm", valueOnly = TRUE)
  y0 <- convertY(x$y0, "mm", valueOnly = TRUE)
  y1 <- convertY(x$y1, "mm", valueOnly = TRUE)
  amplitude <- convertWidth(x$amplitude, "mm", valueOnly = TRUE)

  # Leave frequency and n untouched
  frequency <- x$frequency
  n <- x$n

  # Transform the input data to a data frame containing wave paths
  waves <- lapply(seq_along(x0), function(i) {
    cbind(
      create_wave(
        x = x0[i],
        y = y0[i],
        xend = x1[i],
        yend = y1[i],
        amplitude = amplitude[i],
        frequency = frequency[i],
        n = n
      ),
      id = i
    )
  })
  waves <- do.call(rbind, waves)

  # Construct the grob
  wave_paths <- polylineGrob(
    x = waves$x,
    y = waves$y,
    id = waves$id,
    default.units = "mm",
    gp = x$gp
  )
  setChildren(x, gList(wave_paths))
}

# The GeomWave
GeomWave <- ggproto("GeomWave", Geom,

  # Check that the user has specified sensible parameters
  setup_params = function(data, params) {
    if (is.null(params$n)) {
      params$n <- 50
    } else if (params$n <= 0) {
      rlang::abort("Waves must be defined with `n` greater than 0")
    }
    params
  },

  # Check input data and return grobs
  draw_panel = function(data,
                        panel_params,
                        coord,
                        n = 50,
                        lineend = "butt",
                        na.rm = FALSE) {
    # Remove missing data, returning early if all are missing
    data <- remove_missing(
      df = data,
      na.rm = na.rm,
      vars = c("x", "y", "xend", "yend", "linetype", "linewidth"),
      name = "geom_wave"
    )
    if (is.null(data) || nrow(data) == 0) {
      return(zeroGrob())
    }

    # Supply the coordinate system for the plot
    if (!coord$is_linear()) {
      rlang::warn(
        "Wave geom only works correctly on linear coordinate systems"
      )
    }
    coord <- coord$transform(data, panel_params)

    # Construct the grob
    waveGrob(
      coord$x,
      coord$y,
      coord$xend,
      coord$yend,
      default.units = "native",
      amplitude = unit(coord$amplitude, "cm"),
      frequency = coord$frequency,
      n = n,
      gp = gpar(
        col = alpha(coord$colour, coord$alpha),
        lwd = coord$linewidth * .pt,
        lty = coord$linetype,
        lineend = lineend
      )
    )
  },

  # Specify the default and required aesthetics
  required_aes = c("x", "y", "xend", "yend"),
  default_aes = aes(
    colour = "black",
    linewidth = 0.5,
    linetype = 1L,
    alpha = NA,
    amplitude = 0.1,
    frequency = 1
  )
)


# Constructor wrapper function around WaveGrob
geom_wave <- function(mapping = NULL,
                      data = NULL,
                      stat = "identity",
                      position = "identity",
                      ...,
                      n = 50,
                      lineend = "butt",
                      na.rm = FALSE,
                      show.legend = NA,
                      inherit.aes = TRUE) {
  layer(
    data = data,
    mapping = mapping,
    stat = stat,
    geom = GeomWave,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(
      n = n,
      lineend = lineend,
      na.rm = na.rm,
      ...
    )
  )
}


# And the stat_* constructor function
# The StatWave
StatWave <- ggproto("StatWave", Stat,
  setup_data = function(data, params) {
    if (anyDuplicated(data$group)) {
      data$group <- paste(data$group, seq_len(nrow(data)), sep = "-")
    }
    data
  },
  compute_panel = function(data, scales, n = 50) {
    cols_to_keep <- setdiff(names(data), c("x", "y", "xend", "yend"))
    waves <- lapply(seq_len(nrow(data)), function(i) {
      wave_path <- create_wave(
        data$x[i],
        data$y[i],
        data$xend[i],
        data$yend[i],
        data$amplitude[i],
        data$frequency[i],
        n
      )
      cbind(wave_path, unclass(data[i, cols_to_keep]))
    })
    do.call(rbind, waves)
  },
  required_aes = c("x", "y", "xend", "yend"),
  optional_aes = c("amplitude", "frequency")
)


# Constructor wrapper function around StatWave
stat_wave <- function(mapping = NULL,
                      data = NULL,
                      geom = "wave",
                      position = "identity",
                      ...,
                      n = 50,
                      na.rm = FALSE,
                      show.legend = NA,
                      inherit.aes = TRUE) {
  layer(
    data = data,
    mapping = mapping,
    stat = StatWave,
    geom = geom,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(
      n = n,
      na.rm = na.rm,
      ...
    )
  )
}


# Scales
scale_amplitude_continuous <- function(..., range = c(0.05, 0.2)) {
  continuous_scale(
    aesthetics = "amplitude",
    scale_name = "amplitude_c",
    palette = scales::rescale_pal(range),
    ...
  )
}
scale_amplitude <- scale_amplitude_continuous
scale_amplitude_discrete <- function(...) {
  rlang::abort("Amplitude cannot be used with discrete data")
}
scale_frequency_continuous <- function(...,
                                       range = c(1, 5),
                                       unit = "cm") {
  continuous_scale(
    aesthetics = "frequency",
    scale_name = "frequency_c",
    palette = scales::rescale_pal(range),
    ...
  )
}
scale_frequency <- scale_frequency_continuous
scale_frequency_discrete <- function(...) {
  rlang::abort("Frequency cannot be used with discrete data")
}

draw_key_wave <- function(data, params, size) {
  waveGrob(
    x0 = 0,
    y0 = 0,
    x1 = 1,
    y1 = 1,
    amplitude = unit(data$amplitude, "cm"),
    frequency = data$frequency,
    gp = gpar(
      col = alpha(data$colour %||% "black", data$alpha),
      lwd = (data$size %||% 0.5) * .pt,
      lty = data$linetype %||% 1
    ),
    vp = viewport(clip = "on")
  )
}
GeomWave$draw_key <- draw_key_wave
```

### Step 1

First, a function named `create_wave` is defined to generate a wave-like path between two points `(x, y)` and `(xend, yend)` on a plot. The wave's characteristics, such as amplitude, frequency, and number of points (`n`), are taken as input, and validation checks are performed on the parameters. The output is a data frame with transformed `x` and `y` coordinates, where the sine function is applied to create the wave effect.

```{r}
#| eval: false
#| echo: true
#| code-fold: false

# Defining a wave creating function
create_wave <- function(x,
                        y,
                        xend,
                        yend,
                        amplitude = 1,
                        frequency = 1,
                        n = 500) {
  # Validate the input arguments
  if (amplitude < 0) {
    rlang::abort("`amplitude` must be non-negative.")
  }
  if (frequency <= 0) {
    rlang::abort("`frequency` must be larger than zero.")
  }
  if (n == 0) {
    rlang::abort("`n` must be greater than zero.")
  }

  # Calculate the direct length of the wave path
  length <- sqrt((x - xend)^2 + (y - yend)^2)

  # Calculate the number of points we need
  n_points <- n

  # Calculate the sequence of x and y offset values
  x_seq <- seq(x, xend, length.out = n_points)
  y_seq <- seq(y, yend, length.out = n_points)

  # Create and return the transformed data frame
  data.frame(
    x = x_seq,
    y = y_seq + sin((x_seq - x) * frequency / (length * pi * 2)) * amplitude
  )
}


```

### Step 2

A function `waveGrob` is defined to create a graphical object (grob) representing a wave between two coordinates (`x0, y0` and `x1, y1`) using `grid` graphics. Various parameters such as amplitude, frequency, and number of points are included, with default units being set to "npc" (normalized parent coordinates) if not specified. The function returns a `gTree` object, which is assigned the class "wave".

```{r}
#| eval: false
#| echo: true
#| code-fold: false

# A wave Grob
waveGrob <- function(x0 = unit(0, "npc"),
                     y0 = unit(0, "npc"),
                     x1 = unit(1, "npc"),
                     y1 = unit(1, "npc"),
                     amplitude = unit(0.1, "npc"),
                     frequency = 1,
                     n = 500,
                     default.units = "npc",
                     name = NULL,
                     gp = gpar(),
                     vp = NULL) {
  
  # Use the default unit if the user does not specify one
  if (!is.unit(x0)) x0 <- unit(x0, default.units)
  if (!is.unit(x1)) x1 <- unit(x1, default.units)
  if (!is.unit(y0)) y0 <- unit(y0, default.units)
  if (!is.unit(y1)) y1 <- unit(y1, default.units)
  if (!is.unit(amplitude)) amplitude <- unit(amplitude, default.units)

  # Return a gTree of class "wave"
  gTree(
    x0 = x0,
    y0 = y0,
    x1 = x1,
    y1 = y1,
    amplitude = amplitude,
    frequency = frequency,
    n = n,
    name = name,
    gp = gp,
    vp = vp,
    cl = "wave"
  )
}


```

### Step 3

The function `makeContent.wave` is designed to generate the graphical content for a "wave" grob (**gr**aphical **ob**ject) by converting positions (`x0, y0, x1, y1`) and amplitude to absolute units (in millimeters). A wave path is created for each segment using the previously defined `create_wave` function. The results are combined and passed to `polylineGrob` to create the wave paths, which are then set as children of the original grob object using `setChildren`.

```{r}
#| eval: false
#| echo: true
#| code-fold: false

makeContent.wave <- function(x) {
  
  # Convert position and amplitude values to absolute units
  x0 <- convertX(x$x0, "mm", valueOnly = TRUE)
  x1 <- convertX(x$x1, "mm", valueOnly = TRUE)
  y0 <- convertY(x$y0, "mm", valueOnly = TRUE)
  y1 <- convertY(x$y1, "mm", valueOnly = TRUE)
  amplitude <- convertWidth(x$amplitude, "mm", valueOnly = TRUE)

  # Leave frequency and n untouched
  frequency <- x$frequency
  n <- x$n

  # Transform the input data to a data frame containing wave paths
  waves <- lapply(seq_along(x0), function(i) {
    cbind(
      create_wave(
        x = x0[i],
        y = y0[i],
        xend = x1[i],
        yend = y1[i],
        amplitude = amplitude[i],
        frequency = frequency[i],
        n = n
      ),
      id = i
    )
  })
  waves <- do.call(rbind, waves)

  # Construct the grob
  wave_paths <- polylineGrob(
    x = waves$x,
    y = waves$y,
    id = waves$id,
    default.units = "mm",
    gp = x$gp
  )
  setChildren(x, gList(wave_paths))
}

```

### Step 4

The `GeomWave` is defined as a `ggproto` object that extends `ggplot2` to create wave-like geometries. It validates parameters through `setup_params`, ensuring that `n` (number of points) is greater than zero. In the `draw_panel` function, missing data is removed, and the coordinate system is transformed before constructing the wave grob using `waveGrob`. The constructor function `geom_wave` wraps this geometry into a convenient layer for use in `ggplot2`, with default aesthetics like color, amplitude, and frequency provided.

```{r}
#| eval: false
#| echo: true
#| code-fold: false

# The GeomWave
GeomWave <- ggproto("GeomWave", Geom,

  # Check that the user has specified sensible parameters
  setup_params = function(data, params) {
    if (is.null(params$n)) {
      params$n <- 50
    } else if (params$n <= 0) {
      rlang::abort("Waves must be defined with `n` greater than 0")
    }
    params
  },

  # Check input data and return grobs
  draw_panel = function(data,
                        panel_params,
                        coord,
                        n = 50,
                        lineend = "butt",
                        na.rm = FALSE) {
    # Remove missing data, returning early if all are missing
    data <- remove_missing(
      df = data,
      na.rm = na.rm,
      vars = c("x", "y", "xend", "yend", "linetype", "linewidth"),
      name = "geom_wave"
    )
    if (is.null(data) || nrow(data) == 0) {
      return(zeroGrob())
    }

    # Supply the coordinate system for the plot
    if (!coord$is_linear()) {
      rlang::warn(
        "Wave geom only works correctly on linear coordinate systems"
      )
    }
    coord <- coord$transform(data, panel_params)

    # Construct the grob
    waveGrob(
      coord$x,
      coord$y,
      coord$xend,
      coord$yend,
      default.units = "native",
      amplitude = unit(coord$amplitude, "cm"),
      frequency = coord$frequency,
      n = n,
      gp = gpar(
        col = alpha(coord$colour, coord$alpha),
        lwd = coord$linewidth * .pt,
        lty = coord$linetype,
        lineend = lineend
      )
    )
  },

  # Specify the default and required aesthetics
  required_aes = c("x", "y", "xend", "yend"),
  default_aes = aes(
    colour = "black",
    linewidth = 0.5,
    linetype = 1L,
    alpha = NA,
    amplitude = 0.1,
    frequency = 1
  )
)


# Constructor wrapper function around WaveGrob
geom_wave <- function(mapping = NULL,
                      data = NULL,
                      stat = "identity",
                      position = "identity",
                      ...,
                      n = 50,
                      lineend = "butt",
                      na.rm = FALSE,
                      show.legend = NA,
                      inherit.aes = TRUE) {
  layer(
    data = data,
    mapping = mapping,
    stat = stat,
    geom = GeomWave,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(
      n = n,
      lineend = lineend,
      na.rm = na.rm,
      ...
    )
  )
}


```

### Step 5

The `StatWave` is defined as a `ggproto` object to compute the wave paths for each row of data in the `ggplot2` layer. In the `setup_data` function, groups are ensured to be unique by appending row numbers, while the `compute_panel` function generates the wave path using the `create_wave` function for each data point. The `stat_wave` constructor function provides a wrapper around `StatWave` for use in `ggplot2`, allowing wave geometries to be added as a statistical transformation, with options for amplitude, frequency, and the number of points (`n`). This complements the **geom_wave()** and allows use of **StatWave** with other geoms such as `geom_point()` as shown later in @fig-stat.

```{r}
#| eval: false
#| echo: true
#| code-fold: false


# And the stat_* constructor function
# The StatWave
StatWave <- ggproto("StatWave", Stat,
  setup_data = function(data, params) {
    if (anyDuplicated(data$group)) {
      data$group <- paste(data$group, seq_len(nrow(data)), sep = "-")
    }
    data
  },
  compute_panel = function(data, scales, n = 50) {
    cols_to_keep <- setdiff(names(data), c("x", "y", "xend", "yend"))
    waves <- lapply(seq_len(nrow(data)), function(i) {
      wave_path <- create_wave(
        data$x[i],
        data$y[i],
        data$xend[i],
        data$yend[i],
        data$amplitude[i],
        data$frequency[i],
        n
      )
      cbind(wave_path, unclass(data[i, cols_to_keep]))
    })
    do.call(rbind, waves)
  },
  required_aes = c("x", "y", "xend", "yend"),
  optional_aes = c("amplitude", "frequency")
)


# Constructor wrapper function around StatWave
stat_wave <- function(mapping = NULL,
                      data = NULL,
                      geom = "wave",
                      position = "identity",
                      ...,
                      n = 50,
                      na.rm = FALSE,
                      show.legend = NA,
                      inherit.aes = TRUE) {
  layer(
    data = data,
    mapping = mapping,
    stat = StatWave,
    geom = geom,
    position = position,
    show.legend = show.legend,
    inherit.aes = inherit.aes,
    params = list(
      n = n,
      na.rm = na.rm,
      ...
    )
  )
}

```

### Step 6

Several scale functions are defined to handle the mapping of `amplitude` and `frequency` to continuous data in `ggplot2`. The `scale_amplitude_continuous` and `scale_frequency_continuous` functions create continuous scales for amplitude and frequency, rescaling the values within specified ranges. Discrete scales for both amplitude and frequency are restricted, and an error is raised if attempted, indicating these aesthetics cannot be used with discrete data.

```{r}
#| eval: false
#| echo: true
#| code-fold: false


# Scales
scale_amplitude_continuous <- function(..., range = c(0.05, 0.2)) {
  continuous_scale(
    aesthetics = "amplitude",
    scale_name = "amplitude_c",
    palette = scales::rescale_pal(range),
    ...
  )
}
scale_amplitude <- scale_amplitude_continuous
scale_amplitude_discrete <- function(...) {
  rlang::abort("Amplitude cannot be used with discrete data")
}
scale_frequency_continuous <- function(...,
                                       range = c(1, 5),
                                       unit = "cm") {
  continuous_scale(
    aesthetics = "frequency",
    scale_name = "frequency_c",
    palette = scales::rescale_pal(range),
    ...
  )
}
scale_frequency <- scale_frequency_continuous
scale_frequency_discrete <- function(...) {
  rlang::abort("Frequency cannot be used with discrete data")
}


```

### Step 7

A function `draw_key_wave` is defined to create a key (legend) representation for the `GeomWave` geometry. It utilizes the `waveGrob` function to draw a wave in the legend, with parameters such as amplitude, frequency, color, line width, and line type being set based on the provided `data`. The function is then assigned to the `GeomWave$draw_key` property to define how wave keys will be rendered in the plot's legend.

```{r}
#| eval: false
#| echo: true
#| code-fold: false


draw_key_wave <- function(data, params, size) {
  waveGrob(
    x0 = 0,
    y0 = 0,
    x1 = 1,
    y1 = 1,
    amplitude = unit(data$amplitude, "cm"),
    frequency = data$frequency,
    gp = gpar(
      col = alpha(data$colour %||% "black", data$alpha),
      lwd = (data$size %||% 0.5) * .pt,
      lty = data$linetype %||% 1
    ),
    vp = viewport(clip = "on")
  )
}
GeomWave$draw_key <- draw_key_wave
```

## Demonstration of the newly created extension

The following code creates a sample dataset `df` with 10 observations, each representing a wave with specific amplitude and frequency. The `group` column distinguishes between two groups, A and B.

```{r}
#| label: tbl-1
#| tbl-cap: "Raw data for the geom_wave() created randomly"

set.seed(42)

# Create a data frame with the diversing directions (angles)
waves_data <- tibble(
  # All waves start at origin (0, 0)
  x = 0, 
  y = 0,

  # 10 angles, evenly spaced around the circle
  angle = seq(0, pi / 2, length.out = 10),

  # Set a constant length for arrows (or vary if desired)
  length = 1
) |>
  mutate(
    # Generate the xend and yend coordinates
    x_end = length * cos(angle),
    y_end = length * sin(angle),
    amplitude = sample(1:10, replace = F),
    frequency = sample(1:10, replace = F),
    group = rep(c("A", "B"), each = 5)
  ) |>
  select(-length, -angle)

waves_data |> 
  gt() |> 
  tab_header(
    title = "Raw Data for the geom_wave() extension"
  ) |> 
  fmt_number(decimals = 0) |> 
  fmt_number(
    columns = c(x_end, y_end),
    decimals = 2
  ) |> 
  cols_width(
    everything() ~ px(100)
  ) |> 
  gtExtras::gt_theme_nytimes() |> 
  cols_align(align = "center")
```

The following ggplot call maps the `amplitude` and `frequency` columns to aesthetics, allowing the waves to vary in amplitude and frequency. The `colour` aesthetic maps the `group` column to different colors. The resulting plot displays 5 waves in two different colors, demonstrating how `geom_wave` and `stat_wave` can be used to visualize wave-like patterns with varying amplitude and frequency. You can adjust the `range` argument in `scale_amplitude_continuous` and `scale_frequency_continuous` to change the scaling of the amplitude and frequency aesthetics.

```{r}
#| label: fig-2
#| fig-cap: "Demonstrating the geom_wave()"
#| fig-asp: 1.5

waves_data |>
  ggplot(
    aes(
      x = x, y = y, xend = x_end, yend = y_end,
      amplitude = amplitude, frequency = frequency,
      colour = group
    )
  ) +
  geom_wave() +
  scale_amplitude_continuous(range = c(0.4, 0.8)) +
  scale_frequency_continuous(range = c(1, 1000)) +
  scale_color_manual(values = c("blue", "red")) +
  theme_grey() +
  coord_fixed(clip = "off") +
  labs(
    title = "Waves with varying amplitude and frequency",
    subtitle = "Using the geom_wave() with amplitude and frequency as aesthetics",
    x = "X-Variable",
    y = "Y-Variable"
  ) +
  theme(
    legend.key.size = unit(10, "mm"),
    legend.direction = "horizontal",
    legend.box = "vertical",
    legend.position = "bottom"
  )

```

We could also use other geoms, such as `geom_point()` with StatWave for the same dataset, as shown below in @fig-stat, but as expected, the limitations of using a stat is inability to modulate the aesthetics not listed in geom_point().

```{r}
#| label: fig-stat
#| fig-cap: "Demonstrating the StatWave with geom_point()"

waves_data |>
  ggplot(
    aes(
      x = x, y = y, xend = x_end, yend = y_end,
      amplitude = amplitude, frequency = frequency,
      colour = group
    )
  ) +
  stat_wave(
    geom = "point"
  ) +
  scale_amplitude_continuous(range = c(0.4, 0.8)) +
  scale_frequency_continuous(range = c(1, 1000)) +
  scale_color_manual(values = c("blue", "red")) +
  theme_grey() +
  # coord_fixed(clip = "off") +
  labs(
    title = "Waves with varying amplitude and frequency",
    subtitle = "Using the geom_wave() with amplitude and frequency as aesthetics",
    x = "X-Variable",
    y = "Y-Variable"
  ) +
  theme(
    legend.key.size = unit(10, "mm"),
    legend.direction = "horizontal",
    legend.box = "vertical",
    legend.position = "bottom"
  )
```
