---
title: "Chapter 14"
subtitle: "Scales and guides"
author: "Aditya Dahiya"
date: "2024-03-03"
format: 
  html: 
    code-fold: true
    code-copy: hover
    code-link: true
execute: 
  echo: true
  warning: false
  error: false
  cache: true
filters:
  - social-share
share:
  permalink: "https://aditya-dahiya.github.io/ggplot2book3e/Chapter14.html"
  description: "Solutions Manual (and Beyond) for ggplot2: Elegant Graphics for Data Analysis (3e)"
  twitter: true
  facebook: true
  linkedin: true
  email: true
  mastodon: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

::: {.callout-note appearance="minimal"}
This chapter has no exercises (yet!), so we show some examples to demonstrate the customization of scales.
:::

```{r}
#| label: setup

library(tidyverse)
library(scales)
library(ggnewscale)
library(gt)
library(gtExtras)
library(cowplot)
```

A dummy data set to use for demonstrating scales in ggplot2 is shown in @fig-data

```{r}
#| label: fig-data
#| fig-cap: "Dummy data set created for demonstration of scales in ggplot2"

tb <- tibble(
  v_random_number = sample(0:100, size = 10, replace = FALSE),
  v_continuous = 1:10,
  v_discrete = sample(letters, 10),
  v_expo = (1:10)^4
)

tb |> gt() |> gt_theme_538()
```

## **14.1 Theory of scales and guides**

Scale specification, naming scheme and fundamental scale types are shown in @fig-1.

```{r}
#| label: fig-1
#| fig-cap: "The demonstration of Scale specification, naming scheme and fundamental scale types in ggplot2"

g <- tb |> 
  ggplot(
    aes(
      x = v_continuous,
      y = v_random_number,
      colour = v_discrete
    )
  ) +
  geom_point(size = 4) +
  paletteer::scale_color_paletteer_d(
    "palettetown::croconaw"
  ) +
  labs(
    subtitle = "ggplot2 automatically assigns scale_*_continuous() to X and Y axis in this plot.\nThe colour scale becomes scale_colour_discrete()"
  ) +
  theme_cowplot() +
  theme(
    legend.position = "bottom"
  )
g
```

## **14.2 Scale names**

The scales can be easily names, wither in the main `scale_*_*()` function using the argument `name = "..."` , or, more easily using the `labs()` helper function as shown below in @fig-scale-names

```{r}
#| label: fig-scale-names
#| fig-cap: "Using labs() helper function to add a title to legends and axis"

g +
  labs(
    title = "A nice graph with a title",
    x = "A continuous variable from 1 to 10",
    y = "A number (1 to 100)",
    colour = "A discrete variable",
    subtitle = "The labs() function helps to easily add titles, subtites, legend titles and axes titles."
  )

```

## **14.3 Scale breaks**

The `breaks =` argument to the `scale_*_*()` functions is useful in determining the axis-ticks and lines or the colour levels in the legend to be displayed, as shown below in @fig-breaks

```{r}
#| label: fig-breaks
#| fig-cap: "Using breaks argument in scale_*_* functions to show customized intervals of 2 on x-axis, and, 1000 on y-axis"

g1 <- tb |> 
  ggplot(aes(x = v_continuous,
             y = v_expo,
             colour = v_discrete)) + 
  geom_point(
    size = 6
    ) +
  scale_x_continuous(
    breaks = seq(0, 10, 2),
    name = "Breaks at even numbers"
  ) +
  scale_y_continuous(
    name = "Exponentially increasing numbers\nBreaks at interval of 2000",
    breaks = seq(0, 10000, 2000)
  ) +
  labs(
    title = "Using breaks argument in scale_*_* functions",
    subtitle = "This plot shows customized breaks at interval of 2 on x-axis and 1000 on y-axis"
  ) +
  theme_cowplot() +
  theme(legend.position = "none")
g1
```

## **14.4 Scale limits**

The @fig-limits shows an example of setting scale limits using the `limit =` argument in `scale_*_*()` functions in ggplot2. In this example, Y-axis limited upto 8000; and colour scale limtied to levels between `a` to `h` only.

```{r}
#| label: fig-limits
#| fig-cap: "Setting scale limits using limit argument in scale_*_* functions: Y-axis limited upto 8000; and colour scale limtied to some levels only" 

g2 <- g1 +
  scale_y_continuous(
    limits = c(0, 8000),
    name = "Limits from 0 to 8000"
  ) +
  scale_colour_discrete(
    limits = letters[1:8],
    na.value = "lightgrey",
    name = "Limited to characters between a to h"
  ) +
  labs(
    title = "Setting scale limits",
    subtitle = "Y-axis limited upto 8000; and colour scale limtied to a - h only"
  ) +
  theme(legend.position = "bottom")
g2
```

Further, with the `oob =` argument we can control the behavior of out-of-bounds data, as shown in @fig-oob below to squish to data onto the nearest limits.

```{r}
#| label: fig-oob
#| fig-cap: "Using oob argument to limit Y-axis up-to 8000, but with out of bounds value squished to 8000"

g2 +
  scale_y_continuous(
    limits = c(0, 8000),
    name = "Limits from 0 to 8000; with squish oob",
    oob = squish
  ) +
  labs(
    subtitle = "Y-axis limited upto 8000, but with out of bounds value squished to 8000"
  )
```

## **14.5 Scale guides**

To demonstrate the use of guides() function, we customize the colour legend from @fig-breaks to generate a customized legend in @fig-guides

```{r}
#| label: fig-guides
#| fig-cap: "Using guides() with guide_legend() to customize legend"

g4 <- g1 +
  theme(legend.position = "bottom") +
  labs(
    title = "Customized Legend", 
    colour = "Using guide_legend() to customize legend") +
  guides(
    colour = guide_legend(
      direction = "vertical",
      reverse = TRUE,
      nrow = 2,
      override.aes = list(
        size = 4
      ),
      theme = theme(
        legend.text = element_text(
          face = "bold"
        )
      )
    )
  )
g4
```

## **14.6 Scale transformation**

The continuous scales in @fig-guides can be transformed using the `trans` argument in the `scale_*_*()` function, as shown in @fig-trans below.

```{r}
#| label: fig-trans
#| fig-cap: "Continuous scale transformation into a log2 scale using trans argument in the scale_*_*() function"

g4 + 
  scale_y_continuous(
    trans = "log2",
    name = "Transformed Y-Axis\nwith powers of Two"
  ) +
  labs(subtitle = "Transformed Y-Axis to the powers of 2",
       title = "Transformed Axes")

```

## **14.7 Legend merging and splitting**

The @fig-merged shows us that if we’ve mapped colour to both points and lines, the keys will show both points and lines.

```{r}
#| label: fig-merged
#| fig-cap: "Merged legends in ggplot2 - the default behaviour"
g4 +
  geom_col(
    fill = "transparent"
  ) +
  labs(subtitle = "Merged Legends - default option in ggplot2",
       title = "Legend merging and splitting (1)",
       colour = "Merged Legends")
```

The @fig-split shows us how we can split or hide legends for certain geoms - here we-ve hidden the colour legend for lines / column bars - so the key will show only points (and not lines).

```{r}
#| label: fig-split
#| fig-cap: "Hiding specific legends using show.legend = FALSE in one of the geoms"

g4 +
  geom_col(
    fill = "transparent",
    show.legend = FALSE
  ) +
  labs(subtitle = "Splitting / Hiding Legends using show.legend = FALSE in one of the geoms",
       title = "Legend merging and splitting (2)",
       colour = "Legend only for points")
```

## **14.8 Legend key glyphs**

Using the same graph as the last section, the @fig-glyph demonstrates the use of argument `key_glyph =` within a `geom_*()` function to change the glyph used in the key of the graph.

```{r}
#| label: fig-glyph
#| fig-cap: "The use of argument key_glyph with a geom_*() to change the glyph used in the key"

g4 +
  geom_col(
    fill = "transparent",
    key_glyph = draw_key_path
  ) +
  labs(title = "Changing the glyphs for legends",
       colour = "Customized legend glyph")

```
